Fail Faster

### 2025-03-20 paper/paper.tex
more wibbling in beyond strict

\section{Evaluation}
\label{section:eval}
% \jwc{Initial section here about eval}\tr{todo}
All experiments were run on a 64-bit Linux machine with 264 GB RAM and an Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS. The OCaml version of \name and all OCaml benchmarks
were compiled with 4.14.1+BER MetaOCaml \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flag \texttt{-O3}.
The baseline OCaml generators were written with \bq{} 0.16.
\scalaname{} and all Scala benchmarks
were run on Scala 3.6.3
and OpenJDK 21.0.6, using ScalaCheck version 1.17 as the baseline.
We used \texttt{core\_bench} 0.16 \cn in OCaml and \texttt{jmh} 0.4.7 \cn for generator performance microbenchmarking.

@@

As shown in \figref{fig:scala}, \scalaname achieves an even greater performance 
gain---purely from staging---over ScalaCheck than \name does over \bq. This discrepancy arises from
ScalaCheck's representation of generators as functions of type \texttt{size ->
seed -> Option[A]}. Each generator combinator must construct and then
pattern-match on these \texttt{Option} values, introducing significant boxing
and unboxing overhead at each step.  Staging allows us to eliminate this
overhead, resulting in substantial speedups in aggregate.

@@

\begin{figure}[h]
  % \includegraphics[scale=0.33]{allocs.png}
  \includegraphics[scale=0.45]{time.png}
  \caption{Time to generate a value using each strategy, averaged over 10,000 trials. Performance benefits scale with generation size; lower is better. BQ is \bq; Staged SR is \name using \bq's \texttt{splittable\_random} RNG; Staged CSR is \name using \secref{subsection:faster-rng}'s \csplitmix.}
  \label{fig:time}
\end{figure}

@@

Next, we discuss language-by-language the degree to which the \name{} insights about staging could be used.
\begin{itemize}
  \item \textbf{Racket} is the next target that we intend to test the \name{} methodology on.
  The entire Racket philosophy is intertwined with
  using macros to build small eDSLs like the ones we use to write generators, and so it is a natural fit. Indeed,
  there are at least two \cn{} PBT libraries in Racket.
  However, the racket macros literature, while extensive, does not concern itself with staging for performance purposes.
  For this reason, we --- the authorship team of this paper, devoid of much Racket expertise --- decided to not use Racket as a test case in this paper.

  \item \textbf{Haskell} is home to the original PBT implementation \cn{}, and
  so it is natural to ask why we have not yet applied \name{} to it. The basic
  answer is that the Haskell compiler (GHC) is designed specially to eliminate
  the run-time overhead of monadic abstractions! Indeed, because Haskell is
  pure, GHC can aggressively inline and beta-reduce nearly anything as part of
  its normal optimiation steps.  This means that in many cases, the impact of a
  staged PBT library (and indeed, all staged monadic DSLs) negligible.  For
  complex enough programs however, the heuristics GHC uses can degrade, leading
  to performance overheads for monadic code.  For this reason, Haskell does have
  a multi-stage programming system called Template Haskell \cn{}, which is
  sometimes to build eDSLs in contexts where one does not want to rely on the
  optimizer's heuristics.  In short: it is possible to apply \name{} to Haskell
  QuickCheck, but the benefits would be much less pronounced.

  \item \textbf{Rust} shares some similarities with functional languages, and
  indeed, it is host to a number of property-based testing libraries \cn{}. However Rust does not
  structurally encourage monadic eDSLs --- it does not have a special monadic syntax --- and so
  PBT libraries in Rust ask programmers to define generators directly as \texttt{seed -> value} functions.
  For this reason, \name{} does not seem directly applicable to any of the PBT libraries in Rust.
\end{itemize}

>>>>

\section{Evaluation}
\label{section:eval}
% \jwc{Initial section here about eval}\tr{todo}
All experiments were run on a 64-bit Linux machine with 264 GB RAM and an Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS. \camlname and all OCaml benchmarks
were compiled with 4.14.1+BER MetaOCaml \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flag \texttt{-O3}.
The baseline OCaml generators were written with \bq{} 0.16.
\scalaname{} and all Scala benchmarks
were run on Scala 3.6.3
and OpenJDK 21.0.6, using ScalaCheck version 1.17 as the baseline.
We used \texttt{core\_bench} 0.16 \cn in OCaml and \texttt{jmh} 0.4.7 \cn for generator performance microbenchmarking.

@@

As shown in \figref{fig:scala}, \scalaname achieves an even greater performance 
gain---purely from staging---over ScalaCheck than \camlname does over \bq. This discrepancy arises from
ScalaCheck's representation of generators as functions of type \texttt{size ->
seed -> Option[A]}. Each generator combinator must construct and then
pattern-match on these \texttt{Option} values, introducing significant boxing
and unboxing overhead at each step.  Staging allows us to eliminate this
overhead, resulting in substantial speedups in aggregate.

@@

\begin{figure}[h]
  % \includegraphics[scale=0.33]{allocs.png}
  \includegraphics[scale=0.45]{time.png}
  \caption{Time to generate a value using each strategy, averaged over 10,000 trials. Performance benefits scale with generation size; lower is better. BQ is \bq; Staged SR is \camlname using \bq's \texttt{splittable\_random} RNG; Staged CSR is \camlname using \secref{subsection:faster-rng}'s \csplitmix.}
  \label{fig:time}
\end{figure}

@@

Next, we discuss language-by-language the degree to which the \name{} insights about staging could be used.
\begin{itemize}
  \item \textbf{Racket} is the next target that we intend to test on.
  The entire Racket philosophy is intertwined with
  using macros to build small eDSLs like the ones we use to write generators, and so it is a natural fit. Indeed,
  there are at least two \cn{} PBT libraries in Racket.
  However, the racket macros literature, while extensive, does not concern itself with staging for performance purposes.
  For this reason, we --- the authorship team of this paper, devoid of much Racket expertise --- decided to not use Racket as a test case in this paper.

  \item \textbf{Haskell} is home to the original PBT implementation \cn{}, and
  so it is natural to ask why we have not built a \name{} in it yet. The basic
  answer is that the Haskell compiler (GHC) is designed specially to eliminate
  the run-time overhead of monadic abstractions! Indeed, because Haskell is
  pure, GHC can aggressively inline and beta-reduce nearly anything as part of
  its normal optimiation steps.  This means that in many cases, the impact of a
  staged PBT library (and indeed, all staged monadic DSLs) negligible.  For
  complex enough programs however, the heuristics GHC uses can degrade, leading
  to performance overheads for monadic code.  For this reason, Haskell does have
  a multi-stage programming system called Template Haskell \cn{}, which is
  sometimes to build eDSLs in contexts where one does not want to rely on the
  optimizer's heuristics.  In short: it is possible to build a \name{} version of Haskell
  QuickCheck, but the benefits would be much less pronounced.

  \item \textbf{Rust} shares some similarities with functional languages, and
  indeed, it is host to a number of property-based testing libraries \cn{}. However Rust does not
  structurally encourage monadic eDSLs --- it does not have a special monadic syntax --- and so
  PBT libraries in Rust ask programmers to define generators directly as \texttt{seed -> value} functions.
  For this reason, staging does not seem directly applicable to any of the PBT libraries in Rust.
\end{itemize}
